# McNemar's test {#sec-mcnemar}

The McNemar's test (also known as the paired or matched chi-square) is used to determine if there are differences on a dichotomous dependent variable between two related groups. It can be considered to be similar to the paired-samples t-test, but for a dichotomous rather than a continuous dependent variable. The McNemar's test is used to analyze pretest-posttest study designs (observing categorical outcomes more than once in the same patient), as well as being commonly employed in analyzing matched pairs and case-control studies.

When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Applying hypothesis testing
-   Investigate a change in proportion for paired data using the McNemar's test
-   Interpret the results
:::

## Research question and Hypothesis Testing

We consider the data in *asthma* dataset. The dataset contains data from a survey of 86 children with asthma who attended a camp to learn how to self-manage their asthmatic episodes. The children were asked whether they knew (yes or not) how to manage their asthmatic episodes appropriately at both the start and completion of the camp.

In other words, was a significant change in children's knowledge of asthma management between the beginning and completion of the health camp?

::: {.callout-note icon="false"}
## Null hypothesis and alternative hypothesis

-   $H_0$: There was no change in children's knowledge of asthma management between the beginning and completion of the health camp
-   $H_1$: There was change in children's knowledge of asthma management between the beginning and completion of the health camp
:::

## Packages we need

We need to load the following packages:

```{r}
#| message: false
#| warning: false

library(rstatix)
library(janitor)
library(modelsummary)
library(exact2x2)
library(here)
library(tidyverse)
```

## Preraring the data

We import the data *asthma* in R:

```{r}
#| warning: false

library(readxl)
asthma <- read_excel(here("data", "asthma.xlsx"))
```

```{r}
#| echo: false
#| label: fig-asthma
#| fig-cap: Table with data from "asthma" file.

DT::datatable(
  asthma, extensions = 'Buttons', options = list(
    dom = 'tip',
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

We inspect the data and the type of variables:

```{r}
glimpse(asthma)
```

The dataset *asthma* includes 86 children with asthma (rows) and 2 columns, the character (`<chr>`) `know_begin` and the character (`<chr>`) `know_end`. Therefore, we consider the dichotomous dependent variable asthma knowledge (yes/no) between two time points, `know_begin` and `know_end`.

Both measurements `know_begin` and `know_end` should be converted to factors (`<fct>`) using the `convert_as_factor()` function as follows:

```{r}
asthma <- asthma %>%
  convert_as_factor(know_begin, know_end)

glimpse(asthma)
```

## Contigency table

We can obtain the cross-tabulation table of the two measurements for the children's knowledge of asthma:

```{r}
tb3 <- table(know_begin = asthma$know_begin, know_end = asthma$know_end)
tb3
```

::: callout-important
There is a basic difference between this table and the more common two-way table. In this case, the count represents the **number of pairs**, not the number of individuals.
:::

We want to compare the proportion of children's knowledge of asthma management at the beginning with the proportion of children's knowledge of asthma management at the end. We can create a more informative table using the functions from `{janitor}` package for obtaining total percentages and marginal totals.

::: {#exercise-joins .callout-tip}
## Table with total percentages and marginal totals

::: panel-tabset
## janitor

We can create an informative table using the functions from `{janitor}` package for obtaining total percentages and marginal totals:

```{r}
total_tb2 <- asthma %>%
  tabyl(know_begin, know_end) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("all") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns %>%
  adorn_title

knitr::kable(total_tb2) 
```

## modelsummary

The contingency table using the `datasummary_crosstab()` from the `{modelsummary}` package:

```{r}
modelsummary::datasummary_crosstab(know_begin ~ know_end, 
                     statistic = 1 ~ 1 + N + Percent(), 
                     data = asthma)
```
:::
:::

The proportion of children who knew to manage asthma at the beginning is (6+24)/86= 30/86 = 0.349 or 34.9%. The proportion of children who knew to mange asthma at the end is (29+24)/86 = 53/86 = 0.616 or 61.6%.

::: {.callout-note icon="false"}
## Assumption

The basic assumption of the test is that the sum of the **discordant** cells should be larger than 25 (that is fulfilled in our example).
:::

## Run McNemar's test

Finally, we run the McNemar's test:

::: {#exercise-joins .callout-tip}
## McNemar's test

::: panel-tabset
## Base R

```{r}
mcnemar.test(tb3)
```

## rstatix

```{r}
mcnemar_test(tb3)
```
:::
:::

The proportion of children who knew to manage asthma at the end (61.6%) is significant larger compared with the proportion of children who knew to manage asthma at the beginning (34.9%) (p-value \<0.001).

## Exact binomial test

Exact binomial test for 2x2 table when the sum of the **discordant** cells are less than 25:

```{r}
mcnemar.exact(tb3)
```
