# Normal distribution {#sec-normal}

There are several important probability distributions in statistics. However, the normal distribution might be the most important. This distribution is also referred to as "*Gaussian*" or "*Gauss*" distribution.

```{r}
#| include: false
library(stevemisc)

library(here)
library(tidyverse)
```

::: {.callout-caution icon="false"}
## Learning objectives

-   Understand ...
-   Understand ...
:::

## Packages we need

We need to load the following packages:

```{r}
#| eval: false
library(stevemisc)

library(here)
library(tidyverse)
```

## The shape of a normal distribution

A normal distribution is a symmetric "bell-shaped" probability distribution where most of the observed data are clustered around a central location. Data farther from the central location occur less frequently (@fig-normal).

```{r}
#| warning: false
#| label: fig-normal
#| fig-height: 7.2
#| fig-width: 12.5
#| fig-align: "center"
#| fig-cap: The area underneath a Normal Distribution


x <- seq(-4, 4, length=200)                                                    # <1>
df <- data.frame(x)                                                            # <1>

ggplot(df, aes(x)) +                                                           
  stat_function(fun = dnorm) +                                                 
  scale_x_continuous(breaks = c(-3, -2, -1, 0, 1, 2, 3),                       # <2>
                     labels = expression(-3*sigma, -2*sigma, -1*sigma,         # <2>
                                         mu, 1*sigma, 2*sigma, 3*sigma)) +     # <2>
  labs(x = "Variable",
       y = "Probability density") +
  theme(text = element_text(size = 20))                                    

```

1. Create a sequence of 200 numbers between -4 and +4.
2. Add the standard deviation notation in x-axis.



## The properties of a normal distribution

```{r}
#| warning: false
#| label: fig-normal_auc
#| fig-height: 7.2
#| fig-width: 12.5
#| fig-align: "center"
#| fig-cap: The area underneath a Normal Distribution

normal_dist("#522d80","#00868B") + 
    labs(y = "Probability density",
         x = "Variable") +
scale_x_continuous(breaks = c(-3, -2.58, -1.96, -1, 
                              0, 1, 1.96, 2.58, 3),
                   labels = expression(-3*sigma, -2.58*sigma, -1.96*sigma, -1*sigma, 
                                       mu, 1*sigma, 1.96*sigma, 2.58*sigma, 3*sigma)) +
  theme(text = element_text(size = 20), 
        axis.text.x = element_text(size = 12))

```

The Normal distribution has the properties summarized as follows:

-   Bell shaped and symmetrical around the mean. Shape statistics, skewness and excess kurtosis are zero.

-   The peak of the curve lies above the mean.

-   Any position along the horizontal axis (x-axis) can be expressed as a number of standard deviations from the mean.

-   All three measures of central location mean, median, and mode are the same.

-   **The empirical rule** (also called the "68-95-99 rule"). Much of the area (68%) of the distribution is between -1 $\sigma$ below the mean and +1 $\sigma$ above the mean, the large majority (95%) between -1.96 $\sigma$ below the mean and +1.96 $\sigma$ above the mean (often used as a reference range), and almost all (99%) between -2.58 $\sigma$ below the mean and +2.58 $\sigma$ above the mean. The total area under the curve equals to 1 (or 100%), almost -3 $\sigma$ below the mean and +3 $\sigma$ above the mean.
