# Normal distribution {#sec-normal}

There are several important probability distributions in statistics. However, the normal distribution might be the most important. This distribution is also referred to as "*Gaussian*" or "*Gauss*" distribution.

```{r}
#| include: false
library(stevemisc)

library(here)
library(tidyverse)
```

::: {.callout-caution icon="false"}
## Learning objectives

-   Understand ...
-   Understand ...
:::

## Packages we need

We need to load the following packages:

```{r}
#| eval: false
library(stevemisc)

library(here)
library(tidyverse)
```

## The shape of a normal distribution

A normal distribution is a symmetric "bell-shaped" probability distribution where most of the observed data are clustered around a central location. Data farther from the central location occur less frequently (@fig-normal).


::: {.column-margin}
Normal distribution, technically a probability density function, is a distribution defined by two parameters, mean $\mu$ and variance $\sigma^2$. The mean, $\mu$, is a "location parameter", which defines the central tendency. The variance, $\sigma^2$, is the "scale parameter", which defines the width and height of the distribution. It's formally given as:

$$ f(x)={\frac {1}{\sigma {\sqrt {2\pi }}}}e^{-{\frac {1}{2}}\left({\frac {x-\mu }{\sigma }}\right)^{2}} $$

where $\pi \approx 3.142$ and $e \approx 2.718$.
:::



```{r}
#| warning: false
#| label: fig-normal
#| fig-height: 7.2
#| fig-width: 12.5
#| fig-align: "center"
#| fig-cap: The area underneath a Normal Distribution


x <- seq(-4, 4, length=200)                                                    # <1>
df <- data.frame(x)                                                            # <1>

ggplot(df, aes(x)) +                                                           
  stat_function(fun = dnorm) +                                                 
  scale_x_continuous(breaks = c(-3, -2, -1, 0, 1, 2, 3),                       # <2>
                     labels = expression(-3*sigma, -2*sigma, -1*sigma,         # <2>
                                         mu, 1*sigma, 2*sigma, 3*sigma)) +     # <2>
  labs(x = "Variable",
       y = "Probability density") +
  theme(text = element_text(size = 20))                                    

```

1.  Create a sequence of 200 numbers between -4 and +4.
2.  Add the standard deviation notation in x-axis.

## The properties of a normal distribution

```{r}
#| warning: false
#| label: fig-normal_auc
#| fig-height: 7.2
#| fig-width: 12.5
#| fig-align: "center"
#| fig-cap: The area underneath a Normal Distribution

normal_dist("#522d80","#00868B") + 
  labs(y = "Probability density", 
       x = "Variable") +
  scale_x_continuous(breaks = c(-3, -2.58, -1.96, -1, 
                              0, 1, 1.96, 2.58, 3),
                   labels = expression(-3*sigma, -2.58*sigma, -1.96*sigma, -1*sigma, 
                                       mu, 1*sigma, 1.96*sigma, 2.58*sigma, 3*sigma)) +
  theme(text = element_text(size = 20), 
        axis.text.x = element_text(size = 12))

```

The Normal distribution has the properties summarized as follows:

-   Bell shaped and symmetrical around the mean. Shape statistics, skewness and excess kurtosis are zero.

-   The peak of the curve lies above the mean.

-   Any position along the horizontal axis (x-axis) can be expressed as a number of standard deviations from the mean.

-   All three measures of central location mean, median, and mode are the same.

-   **The empirical rule** (also called the "68-95-99 rule"). Much of the area (68%) of the distribution is between -1 $\sigma$ below the mean and +1 $\sigma$ above the mean, the large majority (95%) between -1.96 $\sigma$ below the mean and +1.96 $\sigma$ above the mean (often used as a reference range), and almost all (99%) between -2.58 $\sigma$ below the mean and +2.58 $\sigma$ above the mean. The total area under the curve equals to 1 (or 100%), almost -3 $\sigma$ below the mean and +3 $\sigma$ above the mean.

 

### Shape statistics and normality

There are two shape statistics that can indicate deviation from normality: **skewness** and **kurtosis**.

**A. Skewness**

Skewness is usually described as a measure of a distribution's symmetry -- or lack of symmetry. Skewness values that are **negative** indicate a tail to the **left** (@fig-skew2 a), **zero** value indicate a **symmetric** distribution (@fig-skew2 b), while values that are **positive** indicate a tail to the **right** (@fig-skew2 c).

Skewness values between −1 and +1 indicate an approximate bell-shaped curve. Values from −1 to −3 or from +1 to +3 indicate that the distribution is tending away from a bell shape with \>1 indicating moderate skewness and \>2 indicating severe skewness. Any values **above +3** or **below−3** are a good indication that the variable is **not normally** distributed.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-skew2
#| fig-cap: Types of distribution according to the summetry.
#| fig-subcap: 
#|   - "Left skewed distribution (negatively skewed). The mean and the meadian are too left to the mode."
#|   - "Symmetric distribution (zero skewness). The mean, median and mode are the same."
#|   - "Right skewed distribution (positively skewed). The mean and median are to the right of the mode."
#| layout-ncol: 3

# create a data frame
x <- seq(0, 1, length=200)
y1 <- dbeta(x, 7, 2)
y2 <- dbeta(x, 7, 7)
y3 <- dbeta(x, 2, 7)
df1 <- data.frame(x, y1, y2, y3)

# left skewed distribution
ggplot(df1, aes(x, y1)) +
  geom_line(color="green", linewidth = 1.0) +
  geom_segment(aes(x = 0.7, y = 0, xend = 0.7,  yend = 1.98),
               color = "black", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.67, y = 2.1, label = 'mean', size = 5, color = "black") +
  geom_segment(aes(x = 0.78, y = 0, xend = 0.78,  yend = 2.78), 
               color = "blue", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.75, y = 2.9, label = 'median', size = 5, color = "blue") +
  geom_segment(aes(x = 0.86, y = 0, xend = 0.86,  yend = 3.17), 
               color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.81, y = 3.13, label = 'mode', size = 5, color = "orange") +
  theme_minimal(base_size = 16) +
  coord_cartesian(expand = FALSE, xlim = c(0, NA), ylim = c(0, NA)) +
  labs(title = "Left skewed distribution",
       x = "Variable",
       y = "Probability density")

# symmetric distribution
ggplot(df1, aes(x, y2)) +
  geom_line(color="green", linewidth = 1.0) +
  geom_segment(aes(x = 0.49, y = 0, xend = 0.49,  yend = 2.89), 
               color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.40, y = 2.85, label = 'mode', size = 5, color = "orange") +
  geom_segment(aes(x = 0.5, y = 0, xend = 0.5,  yend = 2.9), 
               color = "black", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.5, y = 2.92, label = 'mean', size = 5, color = "black") +
  geom_segment(aes(x = 0.51, y = 0, xend = 0.51,  yend = 2.89), 
               color = "blue", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.62, y = 2.85, label = 'median', size = 5, color = "blue") +
  theme_minimal(base_size = 16) +
  coord_cartesian(expand = FALSE, xlim = c(0, NA), ylim = c(0, NA)) +
  labs(title = "Symmetric distribution",
       x = "Variable",
       y = "Probability density")

# right skewed distribution
ggplot(df1, aes(x, y3)) +
  geom_line(color="green", linewidth = 1.0) +
  geom_segment(aes(x = 0.15, y = 0, xend = 0.15,  yend = 3.17), 
               color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.22, y = 3.15, label = 'mode', size = 5, color = "orange") +
  geom_segment(aes(x = 0.25, y = 0, xend = 0.25,  yend = 2.5), 
               color = "blue", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.3, y = 2.55, label = 'median', size = 5, color = "blue") +
  geom_segment(aes(x = 0.3, y = 0, xend = 0.3,  yend = 1.9), 
               color = "black", linetype = "dashed", linewidth = 0.8) +
  annotate('text', x = 0.34, y = 1.96, label = 'mean', size = 5, color = "black") +
  theme_minimal(base_size = 16) +
  coord_cartesian(expand = FALSE, xlim = c(0, NA), ylim = c(0, NA)) +
  labs(title = "Right skewed distribution",
       x = "Variable",
       y = "Probability density")
```

**B. Kurtosis**

The other way that distributions can deviate from normality is **kurtosis**. The **excess kurtosis** parameter is a measure of the combined weight of the tails relative to the rest of the distribution. Kurtosis is associated indirect with the peak of the distribution (if the peak of the distribution is too high or too low compared to a "normal" distribution).

Distributions with **negative** excess kurtosis are called ***platykurtic*** (@fig-kurtosis a). If the measure of excess kurtosis is **0** the distribution is **mesokurtic** (@fig-kurtosis b). Finally, distributions with **positive** excess kurtosis are called ***leptokurtic*** (@fig-kurtosis c).

A kurtosis value between −1 and +1 indicates normality and a value between −1 and −3 or between +1 and +3 indicates a tendency away from normality. Values **below −3** or **above +3** strongly indicate **non-normality**.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-kurtosis
#| fig-cap: Types of distribution according to the summetry.
#| fig-subcap: 
#|   - "Platykurtic distribution (negative excess kurtosis)."
#|   - "Mesokurtic distribution (zero excess kurtosis)."
#|   - "Leptokurtic distribution (positive excess kurtosis). "
#| layout-ncol: 3

# create a data frame
x <- seq(-6, 6, length=200)
y1 <- dnorm(x)
y2 <- dnorm(x, sd= 2)
y3 <- dnorm(x, sd= 0.5)
df2 <- data.frame(x, y1, y2, y3)

# platykurtic distribution
ggplot(df2, aes(x, y1)) +
  geom_line(color="black", linewidth = 0.8, linetype = "dashed") +
  geom_line(aes(x, y2), color="deeppink", linewidth = 0.8) +
  annotate('text', x = 1.5, y = 0.3, label = 'normal', size = 5, color = "black") +
  annotate('text', x = 3.5, y = 0.1, label = 'platykurtic', size = 5, color = "deeppink") +
  theme_minimal(base_size = 16) +
  labs(title = "Platykurtic distribution",
       x = "Variable",
       y = "Probability density")

# mesokurtic distribution
ggplot(df2, aes(x, y1)) +
  geom_line(color="deeppink", linewidth = 0.8) +
  annotate('text', x = 1.9, y = 0.31, label = 'mesokurtic', size = 5, color = "deeppink") +
  theme_minimal(base_size = 16) +
  labs(title = "Mesokurtic distribution",
       x = "Variable",
       y = "Probability density")

# leptokurtic distribution
ggplot(df2, aes(x, y1)) +
  geom_line(color="black", linewidth = 0.8, linetype = "dashed") +
  geom_line(aes(x, y3), color="deeppink", linewidth = 0.8) +
  annotate('text', x = 1.7, y = 0.25, label = 'normal', size = 5, color = "black") +
  annotate('text', x = 1.3, y = 0.75, label = 'leptokurtic', size = 5, color = "deeppink") +
  theme_minimal(base_size = 16) +
  labs(title = "Leptokurtic distribution",
       x = "Variable",
       y = "Probability density")
```
