# Receiver Operating Characteristic (ROC) curve {#sec-roc}

When a diagnostic test result is measured in a **continuous scale**, sensitivity and specificity vary with different cut-off points (thresholds). Therefore, a **convenient** cut-off point[^roc-1] must be selected in order to calculate the measures of diagnostic accuracy of the test. **Receiver operating characteristic** (ROC) curve analysis can be used to help with this decision.

[^roc-1]: **NOTE:** there is critique, however, that the binary diagnosis is problematic.

When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Evaluate a diagnostic test
-   Compare different diagnostic tests
-   Interpret the results
:::

## Research question

We want to compare two screening questionnaires for chronic obstructive pulmonary disease (COPD) among smokers aged \>45 years in the primary care setting:

-   *International Primary Care Airways Group (IPAG)* questionnaire (Score: 0-38)
-   *COPD Population Screener (COPDPS)* questionnaire (Score: 0-10)

The diagnosis of COPD was based on spirometric criterion (FEV 1 /FVC \<0.7 following bronchodilation), clinical status (medical history, symptoms and physical examination), and exclusion of other diseases.

## Packages we need

We need to load the following packages:

```{r}
#| message: false

library(pROC)
library(plotROC)
library(epiR)
library(ggsci)
library(here)
library(tidyverse)
```


>**Other relative packages:** OptimalCutpoints, cutpointr


## Preraring the data

We import the data *copd* in R:

```{r}
#| warning: false

library(readxl)
dat <- read_excel(here("data", "copd.xlsx"))
```

```{r}
#| echo: false
#| label: fig-depression
#| fig-cap: Table with data from "copd" file.

DT::datatable(
  dat, extensions = 'Buttons', options = list(
    dom = 'tip',
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

We inspect the data and the type of variables:

```{r}
glimpse(dat)
```

## Using cut-off points and the ROC curve

Based on previous studies, the cut-off points for a positive response are:

-   ≥ 17 for the IPAG questionnaire
-   ≥ 5 for the COPDPS questionnaire.

We can evaluate these cut-off values by calculating their associated measures of diagnostic accuracy (i.e Se, Sp, PPV, NPV).

```{r}
#| warning: false

dat <- dat %>% 
  mutate(IPAG_cat = cut(IPAG, c(0, 17, 38), labels=c("-","+"), 
                        include.lowest = TRUE, right=FALSE),
         COPDPS_cat = cut(COPDPS, c(0, 5, 10), labels=c("-","+"), 
                          include.lowest = TRUE, right=FALSE))

dat <- as.data.frame(dat)

# we need to create a roc object for each questionnaire
roc1 <- roc(dat$diagnosis, dat$IPAG)
roc2 <- roc(dat$diagnosis, dat$COPDPS)
```

::: callout-important
## NOTE

-   For screening purposes such as mammogram, the cut-off point can be selected to favor a higher sensitivity. Thus, a negative test result indicates the absent of the disease (SeNout; sensitive, negative, "rule out" the disease).

-   For confirmative diagnosis purposes, for example, when a chemotherapy is to initiated once the diagnosis is established, the cut-off point can be selected to favor a higher specificity. Thus, a positive test result indicates the presence of disease (SpPin; specificity, positive, "rule in" the disease).
:::

Additionally, for a given diagnostic test, we can consider all cut-off points that give a unique pair of values for sensitivity and specificity. We can plot in a graph, which is known as a **ROC curve**, the sensitivity on the y-axis and 1-specificity (false positives) values on the x-axis for all these possible cut-off points of the diagnostic test. Then, the **area under the ROC curve** (AUC of ROC), also called the c-statistic, can be calculated which is a widely used measure of overall performance.

### IPAG questionnaire

**A. The use of a cut-off value: IPAG score ≥17**

First, we will find the counts of individuals in each of the four possible outcomes in a 2×2 table for the cut-off point of 17:

```{r}
table(dat$IPAG_cat, dat$diagnosis)
```

Next, we reformat the table as follows:

```{r}
tb1 <- as.table(
  rbind(c(203, 644), c(70, 1670))
  )

dimnames(tb1) <- list(
  Test = c("+", "_"),
  Outcome = c("+", "-")
)

tb1
```

```{r}
results_IPAG <- epi.tests(tb1, digits = 3)
results_IPAG
```

The results using the cut-off point of 17 give Se = 0.744 (0.687 - 0.794) and Sp = 0.722 (0.703 - 0.740). We observe that the probability of the absence of COPD given a negative test result is high NPV = 0.960 (95% CI: 0.949, 0.969) in this sample with smokers.

 

**B. The area under the ROC curve of IPAG questionnaire**

First, let's get sensitivity and specificity if a cutoff point of 17 is used for the IPAG questionnaire and calculate the AUC (95% CI) of ROC. We will use these values to provide additional information about the data being displayed:

```{r}
#| warning: false

# get the sensitivity
se_IPAG <- round(results_IPAG$detail$est[3], digits = 3)

# get the specificity
sp_IPAG <- round(results_IPAG$detail$est[4], digits = 3)

# get AUC IPAG and the 95%CI for this area
auc_values1 <- round(ci.auc(roc1), 3)
ci_l1 <- auc_values1[1]
auc1 <- auc_values1[2]
ci_u1 <- auc_values1[3]
```

 

The ability of the IPAG questionnaire to discriminate between individuals with and without COPD is shown graphically by the ROC curve in (@fig-roc1):

```{r}
#| warning: false
#| label: fig-roc1
#| fig-width: 7
#| fig-cap: The ROC curve of IPAG questionnaire.


# create the plot
g1 <- ggplot(dat, aes(d = diagnosis, m = IPAG)) + 
  geom_roc(n.cuts = 0, color = "#0071BF") +
  theme(text = element_text(size = 14)) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  geom_point(aes(x = 1-sp_IPAG, y = se_IPAG), color = "black") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "1 - Specificity", y = "Sensitivity")

# add annotations to the plot
g1 + annotate("text", x=0.70, y=0.30, 
           label=paste("AUC IPAG = ", auc1, 
                       " (95% CI = ", ci_l1, " - ", ci_u1, ")")) +
  annotate("text", x=0.18, y=0.78, 
           label=paste("Se =", se_IPAG)) +
  annotate("text", x=0.18, y=0.74, 
           label=paste("1-Sp =", 1-sp_IPAG))
```

The AUC of IPAG questionnaire equals to 0.799 (95% CI: 0.769 - 0.829) which indicates a reasonable diagnostic test.

 

::: callout-important
## Perfect and useless diagnostic tests

A test which is perfect at discriminating between those with disease and those without disease has an AUC = 1 (i.e. the ROC curve gets to the upper left-hand corner).

The dashed diagonal line connecting (0,0) to (1,1) is the ROC curve corresponding to a test that is completely useless in diagnosis of a disease, AUC = 0.5 (i.e. individuals with and without the disease have equal "chances" of testing positive).
:::

 

### COPDPS questionnaire

**A. The use of a cut-off value: COPDPS score ≥5**

First, we will find the counts of individuals in each of the four possible outcomes in a 2×2 table for the cut-off point of 5:

```{r}
table(dat$COPDPS_cat, dat$diagnosis)
```

Next, we reformat the table as follows:

```{r}
tb2 <- as.table(
  rbind(c(152, 225), c(121, 2089))
  )

dimnames(tb2) <- list(
  Test = c("+", "_"),
  Outcome = c("+", "-")
)

tb2
```

```{r}
results_COPDPS <- epi.tests(tb2, digits = 3)
results_COPDPS
```

The results using the cut-off point of 5 give Se = 0.577 (0.496 - 0.617) and Sp = 0.903 (0.890 - 0.915).

 

**B. The area under the ROC curve of COPDPS questionnaire**

Similarly, let's get sensitivity and specificity if a cutoff point of 5 is used for the COPDPS questionnaire and calculate the AUC (95% CI) of ROC to provide additional information in the plot:

```{r}
#| warning: false

# get the sensitivity
se_COPDPS <- round(results_COPDPS$detail$est[3], digits = 3)

# get the specificity
sp_COPDPS <- round(results_COPDPS$detail$est[4], digits = 3)

# get AUC COPDPS and the 95%CI for this area
auc_values2 <- round(ci.auc(roc2), 3)
ci_l2 <- auc_values2[1]
auc2 <- auc_values2[2]
ci_u2 <- auc_values2[3]
```

The ROC curve of IPAG questionnaire (@fig-roc2) follows:

```{r}
#| warning: false
#| label: fig-roc2
#| fig-width: 7
#| fig-cap: The ROC curve of COPDPS questionnaire.


# create the plot
g2 <- ggplot(dat, aes(d = diagnosis, m = COPDPS)) + 
  geom_roc(n.cuts = 0, color = "#EFC000") +
  theme(text = element_text(size = 14)) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  geom_point(aes(x = 1-sp_COPDPS, y = se_COPDPS), color = "black") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "1 - Specificity", y = "Sensitivity")

# add annotations to the plot
g2 + annotate("text", x=0.70, y=0.25, 
           label= paste("AUC COPDPS = ", auc2, 
                        " (95% CI = ", ci_l2, " - ", ci_u2, ")")) +
  annotate("text", x=0.19, y=0.56, 
           label=paste("Se =", se_COPDPS)) +
  annotate("text", x=0.19, y=0.52, 
           label=paste("1-Sp =", 1-sp_COPDPS))

```

The AUC of COPDPS questionnaire equals to 0.791 (95% CI: 0.760 - 0.821) which is close to the value 0.799 of AUC of IPAG questionnaire.

## Comparing ROC Curves

**A. Graphical comparison of ROC curves**

We can plot the ROC curves for both questionnaires in the same graph and compare the area under the curves (@fig-rocs):

```{r}
#| warning: false
#| label: fig-rocs
#| fig-width: 7
#| fig-cap: Graphical comparison between IPAG and COPDPS ROC curves.

# prepare the data
longdata <- melt_roc(dat, "diagnosis", c("IPAG", "COPDPS"))
longdata$name = factor(longdata$name, levels = c("IPAG", "COPDPS"))

# create the plot
g <- ggplot(longdata, aes(d = D, m = M, color = name)) + 
  geom_roc(n.cuts = 0) +
  theme(text = element_text(size = 14),
        legend.position="top") +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_color_jco() +
  labs(x = "1 - Specificity", y = "Sensitivity", colour="Questionnaire")

# add annotations to the plot
g + annotate("text", x=0.70, y=0.35, color = "#0071BF",
           label=paste("AUC IPAG = ", auc1, 
                       " (95% CI = ", ci_l1, " - ", ci_u1, ")")) +
  annotate("text", x=0.70, y=0.28, color = "#EFC000",
           label= paste("AUC COPDPS = ", auc2, 
                        " (95% CI = ", ci_l2, " - ", ci_u2, ")"))
```

The AUC values obtained from the ROC curve were 0.799 (95% CI: 0.769 - 0.829) for the IPAG questionnaire and 0.791 (95% CI: 0.760 - 0.821) for the COPDPS questionnaire. Therefore, the two questionnaires have similar overall performance in the present sample.

 

**B. Compare AUCs using the DeLong' s test**

The DeLong's test can be used for comparing 2 areas under the curve (AUCs).

::: {.callout-note icon="false"}
## Null hypothesis and alternative hypothesis

-   $H_0$: there is no difference between two AUCs ($AUC_{IPAG} = AUC_{COPDPS}$)
-   $H_1$: there is difference between two AUCs ($AUC_{IPAG} \neq AUC_{COPDPS}$)
:::

```{r}
#| warning: false

roc.test(roc1, roc2, method=c("delong"))

```

There was no significant difference in the AUC values with the two questionnaires (p = 0.45 \< 0.05).
