# Two-sample t-test (Student's t-test) {#sec-two_samples}

Two sample t-test (Student's t-test) can be used if we have two independent (unrelated) groups (e.g., males-females, treatment-non treatment) and one quantitative variable of interest.

When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Applying hypothesis testing
-   Compare two independent samples applying Student's t-test
-   Interpret the results
:::

## Research question and Hypothesis Testing

We consider the data in *depression* dataset. In an experiment designed to test the effectiveness of paroxetine for treating bipolar depression, the participants were randomly assigned into two groups (paroxetine Vs placebo). The researchers used the Hamilton Depression Rating Scale (HDRS) to measure the depression state of the participants and wanted to find out if the HDRS score is different in paroxetine group as compared to placebo group at the end of the experiment. The significance level $\alpha$ was set to 0.05.

**Note:** A score of 0--7 in HDRS is generally accepted to be within the normal range, while a score of 20 or higher indicates at least moderate severity.

::: {.callout-note icon="false"}
## Null hypothesis and alternative hypothesis

-   $H_0$: the means of HDRS in the two groups are equal ($\mu_{1} = \mu_{2}$)
-   $H_1$: the means of HDRS in the two groups are not equal ($\mu_{1} \neq \mu_{2}$)
:::

## Packages we need

We need to load the following packages:

```{r}
#| message: false
#| warning: false

library(rstatix)
library(PupillometryR)
library(gtsummary)
library(ggpubr)
library(sjstats)
library(sjPlot)
library(ggsci)
library(report)
library(here)
library(tidyverse)
```

## Preraring the data

We import the data *depression* in R:

```{r}
#| warning: false

library(readxl)
depression <- read_excel(here("data", "depression.xlsx"))
```

```{r}
#| echo: false
#| label: fig-depression
#| fig-cap: Table with data from "depression" file.

DT::datatable(
  depression, extensions = 'Buttons', options = list(
    dom = 'tip',
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

We inspect the data and the type of variables:

```{r}
glimpse(depression)
```

The data set *depression* has 76 patients (rows) and includes two variables (columns). The numeric (`<dbl>`) `HDRS` variable and the character (`<chr>`) `intervention` variable which should be converted to a factor (`<fct>`) variable using the `factor()` function as follows:

```{r}
depression <- depression %>% 
  mutate(intervention = factor(intervention))
glimpse(depression)
```

## Assumptions

::: {.callout-note icon="false"}
## Check if the following assumptions are satisfied

A.  The data are **normally** distributed in **both** groups
B.  The data in both groups have similar **variance** (also named as homogeneity of variance or homoscedasticity)
:::

**A. Explore the characteristics of distribution for each group and check for normality**

The distributions can be explored visually with appropriate plots. Additionally, summary statistics and significance tests to check for normality (e.g., Shapiro-Wilk test) can be used.

**Graphs**

We can visualize the distribution of HDRS for the two groups:

```{r}
#| warning: false
#| label: fig-violin_plot1
#| fig-cap: Raincloud plot of HDRS variable stratified by intervention.

set.seed(123)
ggplot(depression, aes(x=intervention, y=HDRS)) + 
  geom_flat_violin(aes(fill = intervention), scale = "count") +
  geom_boxplot(width = 0.14, outlier.shape = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.05), 
             size = 1.2, alpha = 0.6) +
  scale_fill_jco() +
  theme_classic(base_size = 14) +  
  labs(title = "Grouped Raincloud Plot: HDRS by intervention") +
  theme(legend.position="none")

```

The above figure shows that the data are close to symmetry and the assumption of a normal distribution is reasonable.

```{r}
#| warning: false
#| label: fig-qq_HDRS
#| fig-cap: Normality Q-Q plot for HDRS for paroxetine and placebo.

depression %>%
  ggqqplot("HDRS", color = "intervention", conf.int = F) +
  scale_color_jco() +
  facet_wrap(~ intervention) + 
  theme(legend.position = "none")
```

 

**Summary statistics**

The HDRS summary statistics for each group are:

::: {#exercise-joins .callout-tip}
## Summary statistics by group

::: panel-tabset
## dplyr

```{r}
HDRS_summary <- depression %>%
  group_by(intervention) %>%
  dplyr::summarise(
    n = n(),
    na = sum(is.na(HDRS)),
    min = min(HDRS, na.rm = TRUE),
    q1 = quantile(HDRS, 0.25, na.rm = TRUE),
    median = quantile(HDRS, 0.5, na.rm = TRUE),
    q3 = quantile(HDRS, 0.75, na.rm = TRUE),
    max = max(HDRS, na.rm = TRUE),
    mean = mean(HDRS, na.rm = TRUE),
    sd = sd(HDRS, na.rm = TRUE),
    skewness = EnvStats::skewness(HDRS, na.rm = TRUE),
    kurtosis= EnvStats::kurtosis(HDRS, na.rm = TRUE)
  ) %>%
  ungroup()

HDRS_summary
```

## dlookr

```{r}
depression %>% 
  group_by(intervention) %>% 
  dlookr::describe(HDRS) %>% 
  select(described_variables,  intervention, n, na, mean, sd, p25, p50, p75, skewness, kurtosis) %>% 
  ungroup()
```
:::
:::

The means are close to medians (20.3 vs 21 and 21.5 vs 21). The skewness is approximately zero (symmetric distribution) and the (excess) kurtosis falls into the acceptable range of \[-1, 1\] indicating approximately normal distributions for both groups.

 

**Normality test**

::: column-margin
**Hypothesis testing for Shapiro-Wilk test for normality** <br> $H_{0}$: the data came from a normally distributed population. <br> $H_{1}$: the data tested are not normally distributed. <br>

-   If p − value \< 0.05, reject the null hypothesis, $H_{0}$. <br>
-   If p − value ≥ 0.05, do not reject the null hypothesis, $H_{0}$.
:::

The *Shapiro-Wilk* test for normality for each group is:

```{r}
depression %>%
  group_by(intervention) %>%
  shapiro_test(HDRS) %>% 
  ungroup()
```

The tests of normality suggest that the data for the `HDRS` in both groups are normally distributed (p=0.67 \>0.05 and p=0.61 \>0.05, respectively).

::: callout-important
**Normality tests often are not helpful guides**

-   For **small** sample sizes, the Shapiro-Wilk test (and other normality tests) has little power to reject the null hypothesis (under-powered test).

-   If the sample size is **large** normality tests may detect even trivial deviations from the normal distribution (over-powered test).
:::

::: callout-tip
The decision about normality of data should be based on a careful consideration of all available information such as graphs (histograms, Q-Q plots), summary and shape measures and statistical tests.
:::

 

**B. Levene's test for equality of variances**

::: column-margin
**Hypothesis testing for Levene's test for equality of variances**<br> $H_{0}$: the variances of data in two groups are equal<br> $H_{1}$: the variances of data in two groups are not equal<br>

-   If p − value \< 0.05, reject the null hypothesis, $H_{0}$ <br>
-   If p − value ≥ 0.05, do not reject the null hypothesis, $H_{0}$
:::

The *Levene's test* for equality of variances is:

```{r}
depression %>% 
  levene_test(HDRS ~ intervention)
```

Since the p-value = 0.676 \>0.05, the null hypothesis that the variances of HDRs in two groups are equal is not rejected.

## Run the t-test

We will perform a pooled variance t-test (Student's t-test) to test the null hypothesis that the mean HDRS score is the same for both groups (paroxetine and placebo).

::: column-margin
The formula of the test is given by the t-statistic as follows:

$$t = \frac{\bar{x}_{1} - \bar{x}_{2}}{s_{p} \cdot \sqrt{\frac{1}{n_{1}} + \frac{1}{n_{2}}}}$$

where $s_{p}$ is an estimate of the pooled standard deviation of the two groups which is calculated by the following equation:

$$s_{p} = \sqrt{\frac{(n_{1}-1)s_{1}^2 + (n_{2}-1)s_{2}^2}{n_{1}+ n_{2}-2}}$$
:::

::: {#exercise-joins .callout-tip}
## Student's t-test

::: panel-tabset
## Base R

```{r}
t.test(HDRS ~ intervention, var.equal = T, data=depression)                     # <1>
```

1.  If we reject the null hypothesis of Levene's test, we have to type **`var.equal = F`** (or type nothing as this is the default), so the Welch's t-test is applied.

## rstatix

```{r}
depression %>% 
    t_test(HDRS ~ intervention, var.equal = T, detailed = T)                    
```
:::
:::

The p-value = 0.16 is greater than 0.05. There is **no evidence** of a significant difference in mean HDRS between the two groups (failed to reject $H_0$). The difference between means (20.33 - 21.49) equals to -1.16 units of the HDRS and note that the 95% confidence interval of the difference in means (-2.78 to 0.47) **includes** the hypothesized null value of 0. Based on these results, there is not evidence that paroxetine is effective as a treatment for bipolar depression.

Note that the paroxetine group ($n_1=33$) has df (degrees of freedom) = 33-1 = 32 and the placebo sample ($n_2= 43$) has df = 43-1 = 42 , so we have df = 32 + 42 = 74 in total. Another way of thinking of this is that the complete sample size is 76, and we have estimated two parameters from the data (the two means), so we have df = 76-2 = 74 .

::: column-margin
**NOTE:** The Student t-test does not have any restrictions on $n_1$ and $n_2$ ---**they can be equal or unequal**. However, equal samples are preferred because this maximizes the power to detect a specified difference.
:::

 

## Present the results

**Summary table**

It is common practice to report the mean (sd) for each group in summary tables.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

depression %>% 
  tbl_summary(
    by = intervention, 
    statistic = HDRS ~ "{mean} ({sd})", 
    digits = list(everything() ~ 1),
    label = list(HDRS ~ "HDRS score"), 
    missing = c("no")) %>% 
    add_difference(test.args = all_tests("t.test") ~ list(var.equal = TRUE),
                   estimate_fun = HDRS ~ function(x) style_sigfig(x, digits = 2),
                   pvalue_fun = function(x) style_pvalue(x, digits = 2)) %>% 
  add_n()
```

**Report the results**

There is also a specific package with the name {report} that may be useful in reporting the results of the t-test:

```{r}
report_results <- t.test(depression$HDRS ~ depression$intervention, var.equal = T) 
report(report_results)
```

  We can use the above information to write up a final report:

::: callout-tip
## Final report

There is not evidence that HDRS score is significantly different in paroxetine group, mean = 20.3 (sd = 3.7), as compared to placebo group, 21.5 (3.4), (mean difference= -1.2 units, 95% CI = -2.8 to 0.47, p = 0.16 \>0.05).
:::
