# Wilcoxon-Mann-Whitney (Mann-Whitney U) test {#sec-wmw}

The Wilcoxon-Mann-Whitney (WMW) test (sometimes called Mann-Whitney U test or Wilcoxon Rank Sum test) is used to compare two independent samples and is often considered the non-parametric alternative to the Student's t-test when there is violation of normality or for small sample sizes. However, if the samples are very small (both smaller than four observations) then statistical significance is impossible.

When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Applying hypothesis testing
-   Compare two independent samples applying Wilcoxon-Mann-Whitney test
-   Interpret the results
:::

## Research question and Hypothesis Testing

We consider the data in *thromboglobulin* dataset that contains the urinary $\beta$ thromboglobulin excretion (pg/ml) measured in 12 non-diabetic patients and 12 diabetic patients. The researchers used $\alpha$ = 0.05 significance level to test if the distribution of urinary $\beta$ thromboglobulin (b_TG) differs in the two groups.

::: {.callout-note icon="false"}
## Null hypothesis and alternative hypothesis

-   $H_0$: the distribution of urinary $\beta$ thromboglobulin is the same in the two groups
-   $H_1$: the distribution of urinary $\beta$ thromboglobulin is different in the two groups
:::

::: column-margin
**NOTE:** The null hypothesis is that the observations from one group do not tend to have higher or lower ranks than observations from the other group. This test does not compare the medians of the data as is commonly thought, it tests the whole distribution. However, if the distributions of the two groups have similar shapes and spreads (i.e., differing only in location), the WMW test can address (in most cases) whether there are differences in the medians between the two groups.(ref.)
:::

## Packages we need

We need to load the following packages:

```{r}
#| message: false
#| warning: false

library(rstatix)
library(exactRankTests)
library(PupillometryR)
library(gtsummary)
library(ggpubr)
library(sjstats)
library(sjPlot)
library(ggsci)
library(report)
library(here)
library(tidyverse)
```

## Preraring the data

We import the data *thromboglobulin* in R:

```{r}
#| warning: false

library(readxl)
tg <- read_excel(here("data", "thromboglobulin.xlsx"))
```

```{r}
#| echo: false
#| label: fig-depression
#| fig-cap: Table with data from "depression" file.

DT::datatable(
  tg, extensions = 'Buttons', options = list(
    dom = 'tip',
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

We inspect the data and the type of variables:

```{r}
glimpse(tg)
```

The data set *tg* has 24 patients (rows) and includes two variables (columns). The numeric (`<dbl>`) `b_TG` variable and the character (`<chr>`) `status` variable which should be converted to a factor (`<fct>`) variable using the `factor()` function as follows:

```{r}
tg <- tg %>% 
  mutate(status = factor(status))
glimpse(tg)
```

## Explore the characteristics of distribution for each group and check for normality

The distributions can be explored visually with appropriate plots. Additionally, summary statistics and significance tests to check for normality (e.g., Shapiro-Wilk test) can be used.

 

**Graph**

We can visualize the distribution of `b_TG` for the two groups:

```{r}
#| warning: false
#| label: fig-raincloud_tg
#| fig-cap: Rain cloud plot.

set.seed(123)
ggplot(tg, aes(x=status, y=b_TG)) + 
  geom_flat_violin(aes(fill = status), scale = "count") +
  geom_boxplot(width = 0.14, outlier.shape = NA, alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.05), 
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_jco() +
  theme_classic(base_size = 14) +
  theme(legend.position="none", 
        axis.text = element_text(size = 14))
```

The above figure shows that the data in both groups are positively skewed and they have similar shaped distributions.

```{r}
#| warning: false
#| label: fig-qq_tg
#| fig-cap: Normality Q-Q plot for HDRS for paroxetine and placebo.

tg %>%
  ggqqplot("b_TG", color = "status", conf.int = F) +
  scale_color_jco() +
  facet_wrap(~ status) + 
  theme(legend.position = "none")
```

 

**Summary statistics**

The `b_TG` summary statistics for each group are:

::: {#exercise-joins .callout-tip}
## Summary statistics by group

::: panel-tabset
## dplyr

```{r}

tg_summary <- tg %>%
  group_by(status) %>%
  dplyr::summarise(
    n = n(),
    na = sum(is.na(b_TG)),
    min = min(b_TG, na.rm = TRUE),
    q1 = quantile(b_TG, 0.25, na.rm = TRUE),
    median = quantile(b_TG, 0.5, na.rm = TRUE),
    q3 = quantile(b_TG, 0.75, na.rm = TRUE),
    max = max(b_TG, na.rm = TRUE),
    mean = mean(b_TG, na.rm = TRUE),
    sd = sd(b_TG, na.rm = TRUE),
    skewness = EnvStats::skewness(b_TG, na.rm = TRUE),
    kurtosis= EnvStats::kurtosis(b_TG, na.rm = TRUE)
  ) %>%
  ungroup()

tg_summary

```

## dlookr

```{r}
tg %>% 
  group_by(status) %>% 
  dlookr::describe(b_TG) %>% 
  select(described_variables,  status, n, na, mean, sd, p25, p50, p75, skewness, kurtosis) %>% 
  ungroup()
```
:::
:::

The means are not very close to the medians (31.8 vs 29.2 and 13.5 vs 11.0). Moreover, both the skewness (1.81) and the (excess) kurtosis (3.47) for the non-diabetic group falls outside of the acceptable range of \[-1, 1\] indicating right-skewed and leptokurtic distribution.

 

**Normality test**

The *Shapiro-Wilk* test for normality for each group is:

```{r}
tg %>%
  group_by(status) %>%
  shapiro_test(b_TG) %>% 
  ungroup()
```

We can see that the data for the non-diabetic group is not normally distributed (p=0.015 \<0.05) according to the Shapiro-Wilk test.

 

## Run the Wilcoxon-Mann-Whitney test

The difference in location between two distributions with similar shapes (@fig-raincloud_tg) can be tested using the Wilcoxon-Mann-Whitney (WMW) test:

::: callout-tip
## Wilcoxon-Mann-Whitney test

::: panel-tabset
## Base R

```{r}
wilcox.test(b_TG ~ status, conf.int = T, data = tg)
```

**Historical Note:** As you can see, in R the Mann-Whitney test is calculated with the `wilcox.test()` function and it is called Wilcoxon rank-sum test. What is the reason for this? Henry Mann and Donald Whitney (1947) reported in their article that the test was first proposed by Frank Wilcoxon (1945) and they gave their version for the test. So the right would be to call this test Wilcoxon-Mann-Whitney (WMW) test.

## exactRankTests

Although a small number of ties should not have a serious impact on our results, in case of ties we can use the `wilcox.exact()` function from the package {exactRankTests}:

```{r}
wilcox.exact(b_TG ~ status, conf.int = T, data = tg)
```

## rstatix

```{r}
tg %>%
  wilcox_test(b_TG ~ status, detailed = T)
```
:::
:::

The result (the median of the difference[^wmw_test-1] = 18.95, 95%CI: 13.7 to 24) is significant (p \<0.001) and we reject the null hypothesis.

[^wmw_test-1]: **Note:** the estimator for the difference in location parameters does not estimate the difference in medians (a common misconception) but rather the median of the difference between the two samples.

In general, however, WMW test is regarded as a test of comparing the difference in ranks between the two groups as follows:

```{r}
mwu(tg, b_TG, status, out = "browser")
```

 

## Present the results

**Summary table**

It is common practice to report the median (IQR) for each group in summary tables.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

tg %>% 
  tbl_summary(
    by = status, 
    statistic = b_TG ~ "{median} ({p25}, {p75})", 
    digits = list(everything() ~ 1),
    label = list(b_TG ~ "b_TG \n(pg/ml)"), 
    missing = c("no")) %>% 
  add_p(test = b_TG ~ "wilcox.test") %>% 
  add_n()
```

**Report the results**

There is also a specific package with the name {report} that may be useful in reporting the results of WMW test:

```{r}
report_results <- wilcox.test(tg$b_TG ~ tg$status) 
report(report_results)
```

 

We can use the information to write up a final report:

::: callout-tip
## Final report

There is evidence that the urinary $\beta$ thromboglobulin excretion is higher in diabetic group, median = 29.2 (IQR: 27.1, 34.8) pg/ml, as compared to non-diabetic group, 10.9 (8.3, 14.8) pg/ml. The WMW test suggests that there is a significant difference in mean ranks between the two groups (17.67 Vs 7.33, p \<0.001).
:::
